Do professional chess players draw too often?

### 2022-02-18 paper/paper.tex
more of paper

For each game, we modelled $p=(p_{\text{win}}, p_{\text{draw}}, p_{\text{lose}})$
using the following generative model, where $l_w, l_b$ are the latent variables
 computed from the player model for the corresponding players.
\begin{align*}
    p = A l_w + B l_b + \epsilon \\
    \epsilon \sim \mathcal{N}(\mathbf{0}, \sigma\mathbf{I}) \\
    \sigma\sim \Gamma(\alpha=1,\beta=2)
\end{align*}

\section{Results}

>>>>

For each game, we modelled $p=(p_{\text{win}}, p_{\text{draw}}, p_{\text{lose}})$
using the following Bayesian generative model, where $l_w, l_b$ are the latent variables
 computed from the player model for the corresponding players, and $r$ is the result of the game.
\begin{align*}
    \sigma &\sim \Gamma(\alpha=1,\beta=2) \\
    \epsilon  &\sim \mathcal{N}(\mathbf{0}, \sigma\mathbf{I}) \\
    p  &= A l_w + B l_b + \epsilon \\
    p_{\text{pos}}  &= \text{Softplus}(p)= \log(1+\exp(p)) \\
    p_{\text{norm}}  &= \frac{p_{\text{pos}}}{\sum_{i=1}^3 p_{\text{pos}_i}} \\
    r  &\sim \text{Cat}(3, p)
\end{align*}

The result $r$ is drawn from a categorical distribution parameterized by $p_{\text{norm}}$, which in turn is a normalized linear transformation of the latent variables $l_w, l_b$ representing the players.
The parameters are optimized via stochastic variational inference using a family of multivariate Gaussians
as the variational distribution.
I will also try MAP estimation and see the effect on the parameters.

Once matrices $A$ and $B$ are found, we can compute $p_{\text{norm}}$ for each game from the latent variables.
This can then be transformed into $\hat{E}_W$ and $\hat{E}_B$.
Similarly, we can compute $E_W, E_B$ from player Elo ratings, as described previously.

\section{Results}
As of now I did not have enough time to train the model on the full dataset (or even a significant subset).

However, to interpret/compare the results, here a few ideas (a couple of which are quickly explored in the notebooks, but mostly unimplemented):
\begin{itemize}
    \item Compare $\sum_{i=1}^{ngames}E_{W_i}$,
        $\sum_{i=1}^{ngames}\hat{E}_{W_i}$, and
        $\sum_{i=1}^{ngames}r_i$ where $r_j$ is the result of game $j$, and $r=1$ if $W$ wins and $r=0.5$ if it was a draw.
        This is not a very informative measure, but should express whether the expected scores over- or under-estimate the actual scores.
    \item Rank players by average draw probability to form an analogous index, compare with Smerdon's index
    \item Check to see if $\hat{E}_{W_i}$ predicts upsets(victories against players with significantly higher ratings), and if it does so correctly, investigate why
    \item Visualize variance in latent variables as well as in the generated features
    \item Compare/regress draw probabilities across different ratings and/or different differences in rating
    \item Split dataset into training and testing, see if it correctly predicts draws on the test dataset
\end{itemize}

These (future) results can relate back to the article/real-world situation in a couple of ways.
First, they will show whether or not high draw rates are as a result of sub-optimal play rather than higher accuracy at high levels of chess.
This would indicate that other pressures induce players into actually playing worse chess to avoid variance.
This would provide a new angle from which to attack the problem of increasing draws.

The article mentions measures designed to increase errors and decrease the likelihood of draws.
These would likely be successful at the highest echelons of play, where games are sufficiently
scrutinized and play is so precise that suboptimal play essentially does not exist.
However, at lower levels of play (still professional, but just among weaker players), the problem is not absolute precision.
Rather, (assuming the results support this), the problem is this variance-reducing behavior.
So from a design perspective, the solution would be to penalize this behavior.
Making draws worth $\frac{1}{3}$ of a point could discourage taking draws in tournaments, for instance.
Removing agreed draws (where players simply agree to a draw rather than playing it out) as a whole could also make this behavior more difficult.
